---
title: "Reproducible Research: Peer Assessment 1"
author: "Gustavo Gialluisi"
date: "April 06, 2015"
output: html_document
keep_md: true
---

# Reproducible Research: Peer Assessment 1
#### *Gustavo Gialluisi*
#### *April 06, 2015*

This assignment is about using knitr to build this report with answers to the following questions:  
- Loading and preprocessing the data  
- What is mean total number of steps taken per day?  
- What is the average daily activity pattern?  
- Imputing missing values  
- Are there differences in activity patterns between weekdays and weekends?  


## Loading and preprocessing the data

Let's first load all required R packages, on silent mode:
```{r warning=FALSE, message=FALSE}
require(plyr)
require(ggplot2)
library(lubridate)
```

The file [**activity.zip**](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) is required here.  
If the csv file does not exist, we need to unzip the zip file:
```{r}
if(!file.exists("activity.csv")){
    unzip ("activity.zip")
}
```

Load data with **read.csv()**:
```{r}
activity <- read.csv("activity.csv")
```

To count NA values, we'll need a column:
```{r}
activity$is_NA <- is.na(activity$steps)
```


Convert date *string* to *datetime*,  
and convert the interval to datetime to assure correct display in graph:
```{r}
activity$date <- as.Date(activity$date)
activity$interval <- parse_date_time(sprintf("%04d", activity$interval), orders="hm")
```


Have a look at the data:
```{r}
str(activity)
```



## What is mean total number of steps taken per day?
  
We need to summarize steps day by day, I'll use plyr's ddply:
```{r}
stepsperday <- ddply(activity, 
                     .(date), 
                     summarize, 
                     totalsteps = sum(steps, na.rm = TRUE))
```
Calculate mean and median, and plot Histogram, adding the mean and median:
```{r}
mn <- mean(stepsperday$totalsteps, na.rm = TRUE)
mn

md <- median(stepsperday$totalsteps, na.rm = TRUE)
md

ggplot(stepsperday, aes(x=totalsteps)) +
    geom_histogram(binwidth=1500, colour="black", fill="white") + 
    geom_vline(aes(xintercept = mn), color="red", linetype="dashed", size = 1) +#MEAN RED
    geom_vline(aes(xintercept = md), color="blue", linetype="dashed", size = 1) + #MEDIAN BLUE
    labs(title = "Total number of steps taken each day", x = "Total Steps / Day")
```
   
Mean: <span style="color:red">**`r format(mn)`**</span>  
Median: <span style="color:blue">**`r md`**</span>  
   
   


## What is the average daily activity pattern?

We need to calculate average of steps on each interval, with ddply. And plot a profile:

```{r}
avgstepsperinterval <- ddply(activity, 
                         .(interval), 
                         summarize, 
                         avg.interval.steps = mean(steps, na.rm = TRUE))

ggplot(avgstepsperinterval, aes(x=interval, y=avg.interval.steps)) +
    geom_line() +
    labs(title = "Average number of steps taken\naveraged across all days", 
         x = "5-minute interval", y = "total steps\naveraged across all days")
```

  
Verify which interval contains tha maximum value of the steps average:
```{r}
theinterval <- avgstepsperinterval[avgstepsperinterval$avg.interval.steps == max(avgstepsperinterval$avg.interval.steps),]

theinterval
```

The 5-minute interval that contains, on average across all the days in the dataset, the maximum number of steps (**`r format(theinterval$avg.interval.steps)`**) is: **`r theinterval$interval`**.  
    
    
## Imputing missing values

```{r}
sum(activity$is_NA)
```

Rows containing `r NA` value: **`r sum(activity$is_NA)`** out of a total of **`r dim(activity)[1]`** rows (**`r round(100*sum(activity$is_NA)/dim(activity)[1], 1)`%** of the rows).  

The strategy for filling in the missing values in the dataset will be replace the 'NA's values with the average of steps calculated for that 5-minute interval.  

I'll do this merging the activity dataset with average of steps for each 5-minute interval dataset calculated above:
```{r}
activity_fi <- merge(activity, avgstepsperinterval, by="interval")
```

And then replacing NA values with its corresponding average:
```{r}
activity_fi[activity_fi$is_NA,]$steps <- as.integer(activity_fi[activity_fi$is_NA,]$avg.interval.steps)
```


To check impact of imputing missing data, let's replicate the analisys done in the first part of this assignment.  
ddply to summarize:
```{r}
stepsperday_fi <- ddply(activity_fi, 
                     .(date), 
                     summarize, 
                     totalsteps = sum(steps, na.rm = TRUE))
```
  

Calculate mean and median, and plot Histogram, adding the mean and median:
```{r}
mn_fi <- mean(stepsperday_fi$totalsteps, na.rm = TRUE)
mn_fi

md_fi <- median(stepsperday_fi$totalsteps, na.rm = TRUE)
md_fi

ggplot(stepsperday_fi, aes(x=totalsteps)) +
    geom_histogram(binwidth=1500, colour="black", fill="white") + 
    geom_vline(aes(xintercept = mn_fi), color="red", linetype="dashed", size = 1) + #MEAN RED
    geom_vline(aes(xintercept = md_fi), color="blue", linetype="dashed", size = 1) + #MEDIAN BLUE
    labs(title = "Total number of steps taken each day (filled-in data)", x = "Total Steps per Day")
```  


With the filled-in dataset now we have a nicer distribution, and this mean and median values:    
Mean: <span style="color:red">**`r format(mn_fi)`**</span>  
Median: <span style="color:blue">**`r md_fi`**</span>  

   
   


## Are there differences in activity patterns between weekdays and weekends?

- Create the weekend factor column:  

```{r}
activity_fi$daytype <- as.factor( ifelse(wday(activity_fi$date) %in% c(1,7),"weekend", "weekday" ) )
```  


- Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):
```{r}
avgstepsperinterval_fi <- ddply(activity_fi, 
                                .(interval, daytype), 
                                summarize, 
                                avg.interval.steps = mean(steps, na.rm = TRUE))


ggplot(avgstepsperinterval_fi, aes(x=interval, y=avg.interval.steps)) +
    facet_grid(daytype ~ .) +
    geom_line() +
    labs(title = "Average number of steps taken\nper 5-minute interval across weekdays and weekends", 
         x = "5-minute interval", y = "total steps\navg. across weekdays and weekends")

```

The plot above indicates differences in activity patterns between weekdays and weekends.